\documentclass{svproc}
\usepackage{url}
\def\UrlFont{\rmfamily}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}
\mainmatter

\title{Closing the Verification Gap: Non-Intrusive Auditing of Lighting Efficiency in Commercial Buildings}
\titlerunning{Closing the Verification Gap: Non-Intrusive Auditing}

\author{Anonymous Author(s)}
\authorrunning{Anon. Author(s)}
\institute{Affiliation Deleted for Double-Blind Review}

\maketitle

\begin{abstract}
A significant \textit{verification gap} exists in sustainable architecture, where the designed energy efficiency of buildings is rarely achieved in practice, leading to substantial energy and financial waste. This gap persists because the prohibitive cost of hardware sub-metering prevents direct performance auditing at scale. This paper presents an impactful application of data science to bridge this gap, introducing a fully unsupervised, two-stage deep learning framework that enables scalable auditing of lighting efficiency using only aggregate electricity meter data. Our methodology is validated on the public ASHRAE Great Energy Predictor III dataset [1], analyzing nearly 500,000 daily load profiles from 1,448 buildings across 16 climate zones.

In the first stage, a 1D Convolutional Autoencoder learns a latent representation of daily energy profiles, enabling robust clustering of operational patterns. This approach yields a Silhouette Score of 0.37, significantly outperforming baseline clustering on raw data (0.35) and revealing three distinct, interpretable lighting control regimes. In the second stage, an N-BEATS model [15] performs blind source separation, decomposing the aggregate signal into a baseline trend and the isolated lighting load. Our model demonstrates superior performance over traditional STL decomposition and Gradient Boosting baselines. Extensive experiments confirm the model's robustness to noise and its ability to generalize across different climates. A practical case study on an educational facility illustrates how the framework can identify specific, actionable energy-saving opportunities without any hardware retrofits. This work provides a key lesson in applying unsupervised deep learning to real-world problems where ground truth is absent, offering a transparent and reproducible methodology for the applied data science community.

\keywords{Applied Data Science, Building Energy Management, Non-Intrusive Load Monitoring, Energy Disaggregation, Deep Learning, Sustainable Design}
\end{abstract}

\input{section_introduction}
\input{section_related_work}
\input{section_methodology}
\input{section_experiments}
\input{section_conclusion}

\end{document}
